Social Participation Ontology: community documentation, enhancements and use examples

### 2015-01-10 paper.tex
feita revisao completa da apresentacao da ops

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\opsi}{O\textsc{ps}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\sparql}{\textsc{s}par\textsc{ql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\newcommand{\iri}{\textsc{iri}}
\newcommand{\uri}{\textsc{uri}}
\newcommand{\urll}{\textsc{url}}
\newcommand{\ngo}{\textsc{ngo}}
\newcommand{\http}{\textsc{http}}
\newcommand{\opa}{\textsc{op}a}
\newcommand{\ocd}{\textsc{ocd}}
\newcommand{\ontologiaa}{\textsc{o}ntologiaa}
\newcommand{\obs}{\textsc{obs}}
\newcommand{\pubby}{\textsc{p}ubby}
\newcommand{\rdf}{\textsc{rdf}}
\newcommand{\mysql}{\textsc{m}y\textsc{sql}}
\newcommand{\aan}{\textsc{aa}}
\newcommand{\cidadedemocratica}{\textsc{c}idade \textsc{d}emocr\'atica}
\newcommand{\participa}{\textsc{p}articipa.br}
\newcommand{\ontop}{\textsc{o}n\textsc{T}op}
\newcommand{\quest}{\textsc{Q}uest}
\newcommand{\obda}{\textsc{obda}}
\newcommand{\pnud}{\textsc{undp}}
\newcommand{\onu}{\textsc{un}}
\newcommand{\vbs}{\textsc{vbs}}
\newcommand{\lod}{\textsc{lod}}
\newcommand{\corais}{\textsc{c}orais}
\newcommand{\serpro}{\textsc{S}erpro}
\newcommand{\python}{\textsc{P}ython}
% triplification of data

%\newcommand{\t}[1]{\texttt{#1}}
\begin{abstract}
    Participatory democracy advances in virtually all governments. South America presents a prominent context with mixed culture and social predisposition. In 2012, civil, academic and governmental parties started elaborating the ``Common Vocabulary of Social Participation'' (\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social}), as a public and online process. By May, 2013, first reference documents were publicized, together with a preliminary \owl\ code,
logos, and a diagram of a general ``public consultation''.
The \corais\ platform kept online records of the process, like discussions and preparation of texts. 
This article exposes this material and proposes an elementary unfolding: the ``Social Participation Ontology'' (\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}). To exhibit this new ontology, these steps were considered: correction of ontological contradictions and \owl\ protocol use errors; completion of \vcps\ \owl\ code; translations and standardizations; enhancements of class names and labels in Portuguese, Spanish and English; a toy expansion of the ontology by further specifying classes; linked data examples regarding dereferencing, a \sparql\ endpoint and participatory instances; use cases by researchers and public managers. Ongoing work involves further adoption of \ops\ by the official Brazilian social participation federal portal, further adoption by civil participative organizations, and linkage to other participative ontologies. \ops\ is being used as an upper ontology, and all classes linked further to \foaf\ and \bfo\ as higher upper ontologies.
\end{abstract}

@@

Concomitantly, electronic government initiatives are flourishing. Favored mainly by the ubiquity of Internet technologies (e.g. \html\ 5, \node, open source browsers) and by the need for renewal of representative democracy practices. 
These initiatives have taken place in various platforms, including usual social networks (e.g. \facebook, \twitter)
 and dedicated clients created by both government and civil society agents~\cite{socMed,pita2010arquitetura,barros2010alem,knowledge}.

A natural challenge arises: how to link all information produced into an unified knowledge base. This is being addressed, at the technology level, by semantic web developments. Endorsed by World Wide Web Consortium (\wwwc), current semantic web technologies include~\cite{Sem1}:
\begin{itemize}
    \item reasoning by means of ontological specifications,
    \item linking data from different sources (e.g. databases),
    \item organization of domain knowledge for coherent consideration.
\end{itemize}

Key among these technologies, ontologies are considered one of the pillars of the Semantic Web. An ontology is a formal specification of a shared conceptualization~\cite{gruber}.
They are used in the semantic web mainly to give meaning to data, useful for datasets available on the web to make them automatically retrievable and linkable with other datasets. The \wwwc\ created the Web Ontology Language (\owl) as a standard to represent ontologies in the web.
The second version of the language is called \owl\ 2 and offers greater expressive power~\cite{owl2}.

@@

\begin{itemize}
    \item The usage of the word ``vocabulary'' can lead to confusion is some situations as \ops\ is an \owl\ ontology (not, for example, a \skos\ vocabulary).
    \item Documentation seems inconsistent when an ontology is repeatedly called a vocabulary.
    \item \opsi\ is, in fact, an ontology, with a vocabulary, a taxonomic organization and properties further relating the terms.
    \item This coherent naming is a prerequisite for for academic acceptance and further formal adoption by government instances, such as the Brazilian Federal Portal of Open Data~\cite{dadosGov} and of Social Participation~\cite{participabr}.
\end{itemize}

Also, The term \emph{common} was dropped when \ops\ was conceived, as the term is redundant for an ontology. The \vcps\ presented other difficulties, such as missing classes, incorrect URI specifications (containing spaces), some logical flaws, and unnecessary out-of-standards restrictions. This were all solved within \ops\ (to the extent authors were able to, of course).

Next section is dedicated to an inspection of the documents delivered by the \vcps: reference textual documents, images, OWL code, blog posts, discussions and \etherpad s. Section~\ref{exp} presents \ops: modifications made from \vcps\ to \ops: class and property names and labels, class restrictions and property axioms. Section~\ref{ospUtil} is dedicated to \ops\ usage: dereferencing, \sparql\ queries, a toy \ops\ expansion, discursive example of usage, and use cases from government, civil society and academic parties. Concluding remarks are stated with future work, in Section~\ref{conc}.

\section{\vcps\ original documentation}
From April to December, 2012, the \vcps\ was first conceived. In the online process, as registered by Corais platform, 66 users interacted, 6 of them were the most active~\cite{metodologia}. Various materials were produced both as activity traces and as reference media. This section is dedicated to these materials.

\subsection{Reference textual documents}\label{refDocs}
The main documents are:
\begin{itemize}
    \item ``Commented methodology''~\cite{metodologia}: this document describes the public process of ontology conception. It is composed by brief inspections of forum topics, pointing both pertinent characteristics of the online collective process and ontological observations (about classes and properties). Considerations are made about tightening relations to the Open Government Partnership (\ogp), an international initiative to foster transparency and open practices in governments worldwide~\cite{OGP}, and the Brazilian formal action plan, as means to achieve ontology usage. 
     There is also a proposal of a systematic study about electronic government initiatives, so that the \vcps\ might be better contextualized. This document ends by proposing an agenda of meetings with academics, entrepreneurs and government parties.
 \item ``Conceptual modeling, version 0.1 (in natural language)''~\cite{conceptualMod}: this document is a description, in common English, of the \vcps\ ontology. The introduction is mainly a collage of the document above~\cite{metodologia}. Both the itemized description of the ontology, and the considerations for its usage, are of great value as reference. Figure~\ref{fig:v1} is heavily influenced by a diagram related to this document and further described in Section~\ref{sec:im}.
\end{itemize}

@@

\begin{figure}[hb]
    \centering
    \includegraphics[width=0.4\textwidth]{figs/logoUnificadoDoPDF}
    \caption{Some of the various logos for the \ops. (a) is a colored text logo proposal; (b) is a figurative logo; (c) is mixture of both ideas. It can be seen that these logos were conceived for the ontology when it was called a \emph{vocabulary}. Community documents reflect this nomenclature, which is being changed with this article and subsequent work.}
    \label{logo}
\end{figure}

\subsection{OWL code of \vcps}\label{owl}

The \owl\ code for the \vcps\ is online~\cite{owlVcps} and deprecated by \ops\ advent. 
The \vcps\ \owl\ code did not contain all relation from Figure~\ref{fig:v1} (or Figure~\ref{fig:diaorig}). This is directly addressed in Section~\ref{exp}, which exposes the implementation of all relations in the \ops, including \vcps\ \owl\ corrections and adjustments to best practices. The complete and correct \ops\ is further contextualized in Section~\ref{ospUtil}.

\subsection{Blog Posts}
The \vcps\ blog~\cite{coraisBlog}
aggregates both important discussions and documents in no more than twenty posts to date.
All \owl\ code, final documents, public consultations, mental map and images are posted in the blog~\cite{coraisBlog}. 
The first post if from July 24, 2012. Last post about \vcps\ is from May 7, 2013. Most blog posts are from the first day (almost half of them). They received more than
twenty commentaries. Two ``out-of-season'' blog posts, one from August 9, 2012 and another from October  22, 2012, separate first day posts from last posts. Both have about ten commentaries.
Last blog posts occurred as a few days burst and a final message, a month after.

@@

\subsection{Discussions and etherpads}
Besides blog registries of collective elaborations, four etherpads
were written~\cite{etherpads} (these are interfaces that allows writing online texts with multiple simultaneous contributors~\cite{etherpads2}):
\begin{itemize}
    \item A pad for important words.
    \item A pad dedicated to a second phase of \vcps\ elaboration, which did not happen yet.
    \item A pad for process documentation. It became the first document described in Section~\ref{refDocs}. 
    \item A pad for both vocabulary specification and ``questions not addressed to in the webinar''.
\end{itemize}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/opsBFO___}
    \caption{Diagram representation of the Ontology of Social Participation (OSP). Arrows with white heads indicates ``is a'' relations (subclass points to superclass). Arrows with black heads indicates property relations from subject to object . All property relation yield existential restrictions, with the exception of the ``has characteristic'' property, that does not yield restriction.}
    \label{fig:v1}
\end{figure*}

@@

For completeness of exposition and deference to \vcps\ community efforts, consideration of upper ontologies is at Appendix~\ref{upper} and the public consultation model depicted in Figure~\ref{fig:consult} is considered in Appendix~\ref{ap:public}.

\subsection{Standardization and exhibition of implemented features}\label{impl}
Without explicit criteria, \vcps\ \uri\ was \url{http://lumii.lv/ontologies/Corais.owl}. \opsi\ \uri\ was chosen to be \url{http://purl.org/socialparticipation/ops} for the following reasons:
\begin{itemize}
    \item This \uri\ is directly related to the ontology name (\ops).
    \item This \uri, also an \urll, is independent from government and other political associations. This is important to coalesce with interested parties: the Brazilian Federal Social Participation Portal\cite{participa}, Brazilian government repository of vocabularies and ontologies~\cite{vocab}, academic groups,  \ngo s, and non-organized civil society.
    \item This \uri, when reached via \http, can be redirected to where current documentation is held, as it is hosted by \url{http://purl.org}.
\end{itemize}

@@

Upper ontologies usage with \ops\ is under development and should be available in a dedicated work, as possibilities should be inspected carefully. Pertinent and already used as upper ontologies for \ops are \foaf~\cite{foaf} (for linking and describing people and things they do) and \bfo~\cite{bfo} (``designed for use in supporting information retrieval, analysis and integration in scientific and other domains'' as stated on their documentation). Properties were not related to upper ontologies as reasonable relations are still being searched for. Table~\ref{ospClasses} exposes relations of classes to upper ontologies.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/opsTax}
    \caption{A taxonomic tree of the Ontology of Social Participation (OSP). This image was rendered inside Protege, with the OWL code in Appendix~\ref{owl:vcps}. Figure~\ref{fig:v1} contains all this information, but this diagram is more standard and might be simpler for the newcommer. Note that the taxonomic tree does not present any information about properties further linking these classes.}
    \label{fig:owlCC}
\end{figure*}

@@

>>>>

\date{\today}% It is always \today, today,
             %  but any date may be explicitly specified
\newcommand{\ops}{\textsc{ops}}
\newcommand{\opsi}{O\textsc{ps}}
\newcommand{\vcps}{\textsc{vcps}}
\newcommand{\owl}{\textsc{owl}}
\newcommand{\sparql}{\textsc{s}par\textsc{ql}}
\newcommand{\bfo}{\textsc{bfo}}
\newcommand{\dbpedia}{\textsc{db}pedia}
\newcommand{\foaf}{\textsc{foaf}}
\newcommand{\ict}{\textsc{ict}}
\newcommand{\html}{\textsc{html}}
\newcommand{\node}{\textsc{n}ode.js}
\newcommand{\facebook}{\textsc{f}acebook}
\newcommand{\twitter}{\textsc{t}witter}
\newcommand{\wwwc}{\textsc{w3c}}
\newcommand{\skos}{\textsc{skos}}
\newcommand{\etherpad}{\textsc{e}therpad}
\newcommand{\ogp}{\textsc{ogp}}
\newcommand{\iri}{\textsc{iri}}
\newcommand{\uri}{\textsc{uri}}
\newcommand{\urii}{U\textsc{ri}}
\newcommand{\urll}{\textsc{url}}
\newcommand{\ngo}{\textsc{ngo}}
\newcommand{\http}{\textsc{http}}
\newcommand{\opa}{\textsc{op}a}
\newcommand{\ocd}{\textsc{ocd}}
\newcommand{\ontologiaa}{\textsc{o}ntologiaa}
\newcommand{\obs}{\textsc{obs}}
\newcommand{\pubby}{\textsc{p}ubby}
\newcommand{\rdf}{\textsc{rdf}}
\newcommand{\mysql}{\textsc{m}y\textsc{sql}}
\newcommand{\aan}{\textsc{aa}}
\newcommand{\cidadedemocratica}{\textsc{c}idade \textsc{d}emocr\'atica}
\newcommand{\participa}{\textsc{p}articipa.br}
\newcommand{\ontop}{\textsc{o}n\textsc{T}op}
\newcommand{\quest}{\textsc{Q}uest}
\newcommand{\obda}{\textsc{obda}}
\newcommand{\pnud}{\textsc{undp}}
\newcommand{\onu}{\textsc{un}}
\newcommand{\vbs}{\textsc{vbs}}
\newcommand{\lod}{\textsc{lod}}
\newcommand{\corais}{\textsc{c}orais}
\newcommand{\serpro}{\textsc{s}erpro}
\newcommand{\python}{\textsc{p}ython}
\newcommand{\protege}{\textsc{p}rot\`eg\`e}
% triplification of data

%\newcommand{\t}[1]{\texttt{#1}}
\begin{abstract}
    Participatory democracy advances in virtually all governments. South America presents a prominent context with mixed culture and social predisposition. In 2012, civil, academic and governmental parties started elaborating the ``Common Vocabulary of Social Participation'' (\vcps\ from the Brazilian name \emph{Vocabul\'ario Comum de Participa\c{c}\~ao Social}), as a public and online process. By May, 2013, first reference documents were publicized, together with a preliminary \owl\ code,
logos, and a diagram of a general ``public consultation''.
The \corais\ platform kept online records of the process, like discussions and preparation of texts. 
This article exposes this material and proposes an elementary unfolding: the ``Social Participation Ontology'' (\ops\ from the Brazilian name \emph{Ontologia de Participa\c{c}\~ao Social}). To exhibit this new ontology, these steps were considered: correction of ontological contradictions and \owl\ protocol use errors; completion of \vcps\ \owl\ code into a preliminary version of the \ops; translations and standardizations; enhancements of class names and labels in Portuguese, Spanish and English; a toy expansion of the ontology by further specifying classes; linked data examples regarding dereferencing, a \sparql\ endpoint and participatory instances; use cases by researchers and public managers. Ongoing work involves further adoption of \ops\ by the official Brazilian federal portal of social participation, further adoption by civil participative organizations, and linkage to other participative ontologies. \ops\ is being used as an upper ontology, and all classes linked further to \foaf\ and \bfo\ as higher upper ontologies.
\end{abstract}

@@

Concomitantly, electronic government initiatives are flourishing, favored mainly by the ubiquity of Internet technologies (e.g. \html\ 5, \node, open source browsers) and by the need for renewal of representative democracy practices. 
These initiatives have taken place in various platforms, including usual social networks (e.g. \facebook, \twitter)
 and dedicated clients created by both government and civil society agents~\cite{socMed,pita2010arquitetura,barros2010alem,knowledge}.

A natural challenge arises: how to link information produced into an unified knowledge base. This is being addressed, at the technology level, by semantic web developments. Endorsed by World Wide Web Consortium (\wwwc), current semantic web technologies include~\cite{Sem1}:
\begin{itemize}
    \item reasoning by means of ontological specifications,
    \item linking data from different sources (e.g. databases),
    \item organization of domain knowledge for coherent consideration.
\end{itemize}

Key among these technologies, ontologies are considered one of the pillars of the semantic web. An ontology is usually defined as a formal specification of a shared conceptualization~\cite{gruber}.
They give meaning to data and are useful for datasets available on the web to make them automatically retrievable and linkable with other datasets. The \wwwc\ created the Web Ontology Language (\owl) as a standard to represent ontologies in the web.
The second version of the language is called \owl\ 2 and offers greater expressive power~\cite{owl2}, but the version is omitted with very few exceptions.

@@

\begin{itemize}
    \item The usage of the word ``vocabulary'' can lead to confusion is some situations as \ops\ is an \owl\ ontology (not, for example, a \skos\ vocabulary).
    \item Documentation seems inconsistent when an ontology is repeatedly called a vocabulary.
    \item \opsi\ is, in fact, an ontology, with a vocabulary, a taxonomic organization and properties further relating the terms.
    \item This coherent naming is a prerequisite for for academic acceptance and further formal adoption by government instances, such as the Brazilian Federal Portal of Open Data~\cite{dadosGov} and the Brazilian Federal Portal of Social Participation~\cite{participa}.
\end{itemize}

Also, the term \emph{common} was dropped when \ops\ was conceived, as the term is redundant for an ontology. The \vcps\ presented other difficulties, such as missing classes, incorrect \uri\ specifications (containing spaces), some logical flaws, and unnecessary out-of-standards restrictions. This were all solved within \ops\ (to the extent authors were able to, of course).

Next section is dedicated to an inspection of the documents delivered by the \vcps: reference textual documents, images, \owl\ code, blog posts, discussions and \etherpad s. Section~\ref{exp} presents \ops: modifications made from \vcps\ to \ops: class and property names and labels, class restrictions and property axioms. Section~\ref{ospUtil} is dedicated to \ops\ usage: dereferencing, \sparql\ queries, a toy \ops\ expansion, discursive example of usage, and use cases from government, civil society and academic parties. Concluding remarks are stated with future work, in Section~\ref{conc}.

\section{\vcps\ original documentation}
From April to December, 2012, \vcps\ was first conceived. In the online process, as registered by \corais\ platform, 66 users interacted, 6 of them were the most active~\cite{metodologia}. Various materials were produced both as activity traces and as reference media. This section is dedicated to these materials.

\subsection{Reference textual documents}\label{refDocs}
The main documents are:
\begin{itemize}
    \item ``Commented methodology''~\cite{metodologia}: this document describes the public process of \vcps\ conception. It is composed by brief inspections of forum topics, pointing both pertinent characteristics of the online collective process and ontological observations (about classes and properties). Considerations are made about tightening relations with the Open Government Partnership (\ogp), an international initiative to foster transparency and open practices in governments worldwide~\cite{OGP}, and the Brazilian formal action plan, as means to achieve ontology usage. 
     There is also a proposal of a systematic study of electronic government initiatives, so that the \vcps\ might be better contextualized. This document ends by proposing an agenda of meetings with academics, entrepreneurs and government parties.
 \item ``Conceptual modeling, version 0.1 (in natural language)''~\cite{conceptualMod}: this document is a description, in common English, of the \vcps. The introduction is mainly a collage of the document above~\cite{metodologia}. Both the itemized description of the ontology, and the considerations for its usage, are of great value as reference. Figure~\ref{fig:v1} is heavily influenced by a diagram related to this document and further described in Section~\ref{sec:im}.
\end{itemize}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.4\textwidth]{figs/logoUnificadoDoPDF}
    \caption{Some of the various logos for of the \vcps. (a) is a colored text logo proposal; (b) is a figurative logo; (c) is mixture of both ideas. It can be seen that these logos were conceived for the ontology when it was called a \emph{vocabulary}. Community documents reflect this nomenclature, which is being changed with this article and subsequent work with \ops\ advent.}
    \label{logo}
\end{figure}

\subsection{OWL code of VCPS}\label{owl}

The \owl\ code of \vcps\ is online~\cite{owlVcps} and deprecated by \ops\ advent. 
The \vcps\ \owl\ code did not contain all relation from Figure~\ref{fig:v1} (or Figure~\ref{fig:diaorig}). This is directly addressed in Section~\ref{exp}, which exposes the implementation of all relations in the \ops, including \vcps\ \owl\ corrections and adjustments to best practices. The complete and correct \ops\ is further contextualized in Section~\ref{ospUtil}.

\subsection{Blog Posts}
The \vcps\ blog
aggregates both important discussions and documents in no more than twenty posts to date.
All \owl\ code, final documents, public consultations, mental map and images are posted in the blog~\cite{coraisBlog}. 
The first post if from July 24, 2012. Last post about \vcps\ is from May 7, 2013. Most blog posts are from the first day (almost half of them). They received more than
twenty commentaries. Two ``out-of-season'' blog posts, one from August 9, 2012 and another from October  22, 2012, separate first day posts from last posts. Both have about ten commentaries.
Last blog posts occurred as a few days burst and a final message, a month after.

@@

\subsection{Discussions and etherpads}
Besides blog registries of collective elaborations, four \etherpad s
were written~\cite{etherpads} (these are interfaces that allows writing online texts with multiple simultaneous contributors~\cite{etherpads2}):
\begin{itemize}
    \item A pad for important words.
    \item A pad dedicated to a second phase of \vcps\ elaboration, which did not happen yet.
    \item A pad for process documentation. It became the first document described in Section~\ref{refDocs}. 
    \item A pad for both vocabulary specification and ``questions not addressed to in the webinar''.
\end{itemize}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/opsBFO___}
    \caption{Diagram representation of the Ontology of Social Participation (\ops). Arrows with white heads indicate ``is a'' relations (subclass points to superclass). Arrows with black heads indicate property relations from subject to object . All property relation yield existential restrictions, with the exception of the ``has characteristic'' property, that does not yield restriction. Upper ontologies \bfo\ and \foaf\ are asserted through color (\bfo) and dashed boxes (\foaf).}
    \label{fig:v1}
\end{figure*}

@@

\subsection{Standardization and exhibition of implemented features}\label{impl}
Without explicit criteria, \vcps\ \uri\ was \url{http://lumii.lv/ontologies/Corais.owl}. \opsi\ \uri\ was chosen to be \url{http://purl.org/socialparticipation/ops} for the following reasons:
\begin{itemize}
    \item This \uri\ is directly related to the ontology name (\ops).
    \item This \uri, also an \urll, is independent from government and other political associations. This is important to coalesce with interested parties: the Brazilian Federal Social Participation Portal\cite{participa}, Brazilian government repository of vocabularies and ontologies~\cite{vocab}, academic groups,  \ngo s, and non-organized civil society.
    \item Derived \uri s, when reached via \http, can be redirected to where current documentation is held, as it is hosted by \url{http://purl.org}.
\end{itemize}

@@

Upper ontologies usage with \ops\ is under development and should receive a dedicated article, as possibilities should be inspected carefully. Pertinent and already used as upper ontologies for \ops\ are \foaf~\cite{foaf} (for linking and describing people and things they do) and \bfo~\cite{bfo} (``designed for use in supporting information retrieval, analysis and integration in scientific and other domains'' as stated on their documentation). Properties were not related to upper ontologies as reasonable relations are still being searched for. Table~\ref{ospClasses} exposes \ops\ classes and respective upper ontologies classes.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figs/opsTax}
    \caption{A taxonomic tree of the Ontology of Social Participation (\ops). This image was rendered inside \protege, with the \owl\ code in Appendix~\ref{owl:vcps}. Figure~\ref{fig:v1} is more informative, but this diagram is more standard and might be simpler for the newcomer. Note that the taxonomic tree does not present any information about properties further linking these classes. The prefixes {\tt snap:} and {\tt span:} are \bfo\ specific.}
    \label{fig:owlCC}
\end{figure*}

@@

  \footnotesize
